### 2014-10-30 contract-language.tex
Edits
edits

In this section, we formalize the contract language of \name, and describe our
contract classification scheme, which analyzes a contract and maps it to the
weakest store consistency level sufficient to satisfy its consistency
requirements.

@@

\begin{figure}
\begin{smathpar}
\stretcharraybig
\begin{array}{lclcl}
\multicolumn{5}{l}{
  {\eff} \in \mathtt{Effect} \qquad
  {\cv} \in \mathtt{Contract} \qquad
  \set{\eff} \in \mathtt{Effect\; Set}
}\\
\EffSoup & \in & \mathtt{EffSoup}	  & \coloneqq & \set{\eff} \\
\visZ, \soZ, &	\in & \mathtt{Relations} & \coloneqq &
  \set{\eff}\times\set{\eff} \\
\sameobjZ		&     &  & \\
{\E} 		& \in & \mathtt{ExecState}  & \coloneqq & \Exec \\
\end{array}
\end{smathpar}

@@

In this section, we formally define what it means for an execution to
satisfy a contract.

@@

Note that the axiomatic definition an execution ($\E$) provides
interpretations for primitive relations (eg: $\visZ$) that occur free
in contract formulas, and also fixes the domain of quantification to
set of all effects ($\EffSoup$) observed during the program execution.
As such, $\E$ is a potential model for any first-order formula ($\cv$)
expressible in our contract language. If $\E$ is indeed a valid model
for $\cv$ (expressed using model-theoretic consequence relation as $\E
\models \cv$), we say that the execution $\E$ satisfied the contract
$\cv$:
\begin{definition}
An axiomatic execution $\E$ is said to satisfy a contract $\cv$ if and
only if $E \models \cv$.
\end{definition}

@@

\item \textbf{Eventually consistency}: Eventually consistent operations can
  be satisfied as long as the client can reach at least one replica. For
  example, \cf{deposit} is an eventually consistent operation; its semantics
  does not require its action to manifest on all replicas before other
  operations in its session are allowed to proceed. While eventually
  consistent data store typically offer \emph{basic} eventual consistency
  with all possible anomalies, we assume that our store provides stronger
  semantics that remain highly-available~\cite{BailisHAT,COPS}; the store
  always exposes a causal cut of the updates. This semantics can be formally
  captured in terms of the following contract definition:

@@

\item \textbf{Causal consistency}: Causally consistent operations
  are required to see a causally consistent snapshot of the object state,
  including the actions performed on the same session.  The latter
  requirement entails that if two operations $o_1$ and $o_2$ from the
  same session are applied to two different replicas $r_1$ and $r_2$,
  the second operation cannot be discharged until the effect of $o_1$ is
  merged with $o_2$ in both $r_1$ and $r_2$. The \cf{getBalance}
  operation requires causal consistency, as it requires the operations
  from the same session to be visible, which cannot be guaranteed under
  eventual consistency. We assume that causality is only tracked through
  operations on the same object; two operations in the same session but
  on different objects are considered causally unrelated under this
  definition. Stores typically avoid tracking causality across objects
  to mitigate overheads when causality tracking is unnecessary. The
  corresponding store semantics is captured by the contract $\ccc$
  defined below:

@@

We model the system as a tuple $\E;\Sigma$, where the execution $\E$
captures the data store's current state of the execution, and
$\Sigma$, the session soup, is the set of concurrent client sessions
interacting with the store. A session is modeled simply as a sequence
of replicated data type operations ($op$), tagged with the consistency
class ($\tau$) of their contracts (as determined by the contract
classification scheme). We assume rewrite relation of form:

@@

\noindent on the system state. The relation captures store's progress
in execution (from $\E$ to $\E'$) due to the application of an
operation ($op$) from one of the sessions in $\Sigma$ to a replica,
generating a new effect ($\eff$). If the resultant execution ($\E'$)
satisfies the store contract ($\cv(\tau)$) of $\tau$ (i.e., $\E
\models \cv(\tau)$), the we say that the store has \emph{enforced} the
contract $\cv(\tau)$ in the execution $\E'$.

With help of the axiomatic model, we now state the soundness of
contract enforcement as following:

@@

The theorem states that if a data store correctly enforces $\scc$,
$\ccc$, and $\ecc$ contracts in all well-formed executions, then the
same store, extended with classification scheme, can enforce all
well-formed \name contracts. The proof of the theorem has been
relegated to our tech report~\cite{techrep} owing to space
constraints \footnote{Tech report also contains operational semantics of our
implementation of the data store, which concretizes the rewrite
relation ($\xrightarrow{}$), along with the proof that it correctly
enforces all \name contracts.}.

>>>>

In this section, we formalize \name's contract language, and describe our
contract classification scheme, which analyzes a contract and maps it to the
weakest store consistency level sufficient to satisfy its consistency
requirements.

@@

\begin{figure}
\begin{smathpar}
\begin{array}{lclcl}
\multicolumn{5}{l}{
  {\eff} \in \mathtt{Effect} \qquad
  {\cv} \in \mathtt{Contract} \qquad
  \set{\eff} \in \mathtt{Effect\; Set}
}\\
\EffSoup & \in & \mathtt{EffSoup}	  & \coloneqq & \set{\eff} \\
\visZ, \soZ, &	\in & \mathtt{Relations} & \coloneqq &
  \set{\eff}\times\set{\eff} \\
\sameobjZ		&     &  & \\
{\E} 		& \in & \mathtt{ExecState}  & \coloneqq & \Exec \\
\end{array}
\end{smathpar}

@@

@@

Note that the axiomatic definition of an execution ($\E$) provides
interpretations for primitive relations (eg: $\visZ$) that occur free
in contract formulas, and also fixes the domain of quantification to
set of all effects ($\EffSoup$) observed during the program execution.
As such, $\E$ is a potential model for any first-order formula ($\cv$)
expressible in our contract language. If $\E$ is indeed a valid model
for $\cv$ (expressed using the model-theoretic consequence relation as $\E
\models \cv$), we say that the execution $\E$ satisfied the contract
$\cv$:
\begin{definition}
An axiomatic execution $\E$ is said to satisfy a contract $\cv$ if and
only if $E \models \cv$.
\end{definition}

@@

\item \textbf{Eventually consistency}: Eventually consistent operations can
  be satisfied as long as the client can reach at least one replica. For
  example, \cf{deposit} is an eventually consistent operation; its semantics
  does not require its action to manifest on all replicas before other
  operations in its session are allowed to proceed. While eventually
  consistent data stores typically offer \emph{basic} eventual consistency
  with all possible anomalies, we assume that our store provides stronger
  semantics that remain highly-available~\cite{BailisHAT,COPS}; the store
  always exposes a causal cut of the updates. This semantics can be formally
  captured in terms of the following contract definition:

@@

\item \textbf{Causal consistency}: Causally consistent operations are
  required to see a causally consistent snapshot of the object state,
  including the actions performed on the same session.  The latter
  requirement entails that if two operations $o_1$ and $o_2$ from the same
  session are applied to two different replicas $r_1$ and $r_2$, the second
  operation cannot be discharged until the effect of $o_1$ is merged with
  $o_2$ in both $r_1$ and $r_2$. The \cf{getBalance} operation requires
  causal consistency, as it requires the operations from the same session to
  be visible, which cannot be guaranteed under just eventual consistency. We
  assume that causality is only tracked through operations on the same
  object; two operations in the same session but on different objects are
  considered causally unrelated under this definition. Stores typically
  avoid tracking causality across objects to mitigate overheads when
  causality tracking is unnecessary. The corresponding store semantics is
  captured by the contract $\ccc$ defined below:

@@

We model the system as a tuple $\E;\Sigma$, where the execution $\E$
captures the data store's current state of the execution, and
$\Sigma$, the session soup, is the set of concurrent client sessions
interacting with the store. A session is modeled simply as a sequence
of replicated data type operations ($op$), tagged with the consistency
class ($\tau$) of their contracts (as determined by the contract
classification scheme). We assume a rewrite relation of form:

@@

\noindent on the system state. The relation captures the progress of
the execution (from $\E$ to $\E'$)  due to the application of an
operation ($op$) from one of the sessions in $\Sigma$ to a replica,
generating a new effect ($\eff$). If the resultant execution ($\E'$)
satisfies the store contract ($\cv(\tau)$) of $\tau$ (i.e., $\E
\models \cv(\tau)$), then we say that the store has \emph{enforced} the
contract $\cv(\tau)$ in the execution $\E'$.

With help of the axiomatic model, we now state the soundness of
contract enforcement as follows:

@@

The theorem states that if a data store correctly enforces $\scc$, $\ccc$,
and $\ecc$ contracts in all well-formed executions, then the same store,
extended with classification scheme, can enforce all well-formed \name
contracts. The proof of the theorem has been relegated to our tech report
and is provided as part of the supplementary
material~\cite{techrep}\footnote{The supplementary material and tech report
  also defines an operational semantics that concretizes the rewrite
  relation ($\xrightarrow{}$), along with the proof that it correctly
  enforces all \name contracts.}.

### 2014-10-30 motivation.tex
edits

Consider how we might implement a highly available bank account on top of an
eventually consistent data store, with the \emph{integrity} constraint that the
balance should never drop below 0. We will begin by implementing the bank
account replicated data type (RDT) in \name, and then describe the mechanisms
to obtain the desired correctness guarantees.

@@

A key novelty	in \name is that it allows the addition of new RDTs to the store,
immensely improving the programmability of data stores. The programmer no
longer has to coerce the application logic to utilize the store provided data
types. In addition, \name treats the convergence semantics of the data type
separately from the consistency properties. The separation of concern allows
the programmer to \emph{operationally} reason about conflict resolution, while
permitting \emph{declarative} reasoning for consistency. As mentioned in the
previous section, RDTs are described in terms of operations on a RDT object,
where each operation is described as a reduction over the state of the object
i.e., the list of effects. Let us assume that the bank account object provides
three operations: \cf{deposit}, \cf{withdraw} and \cf{getBalance}, with the
assumption that the withdraw fails if the account has insufficient balance.
Every operation in \name is of the following type, written in Haskell syntax:

@@

\noindent It takes a list of effects (the \emph{context} for the operation), an
argument and returns a result along with an optional effect. Read-only
operations do not return any effect. If an effect is returned, it is added to
the state of the object at the current replica, and asynchronously sent to
other replicas. The implementation of the bank account operations in \name is
given below:

\begin{codehaskell}
data Acc = Deposit_ Int | Withdraw_ Int | GetBalance_

@@

withdraw :: [Acc] -> Int -> (Bool, Maybe Acc)
withdraw ctxt v =
	if sel1 $ getBalance ctxt () >= v
  then (True, Just $ Withdraw_ v)
	else (False, Nothing)
\end{codehaskell}

@@

\name helps facilitate the mapping of operations to appropriate consistency
levels by letting the programmer declare application-level consistency
constraints as \emph{contracts} that axiomatically specify the set of allowed
executions involving this operation.  In the case of the bank account example,
any execution that does not exhibit the anomalies described in the previous
section is a \emph{well-formed} execution on the bank account object.  By
specifying the set of legal executions for each data type in terms of a trace
of operation invocations on that type, \name\ ensures that all executions over
that type are well-formed.

@@

\noindent Here, $\cureff$ stands for the effect emitted by the \cf{withdraw} operation.
The syntax $a:\rcf{withdraw}$ states that $a$ is an effect which was emitted
by a \cf{withdraw} operation i.e., $\oper{a}{\rcf{withdraw}}$ holds.  The
contract specifies that if the current operation emits an effect $\cureff$,
then for any operation $a$ which was emitted by a \cf{withdraw} operation, it
is the case that $a = \cureff$ or $a$ is visible to $\cureff$, or vice versa.
Any execution on a bank account object that preserves the above contract for a
\cf{withdraw} operation is said to be derived from a correct implementation of
\cf{withdraw}.

@@

Our specification language (Fig~\ref{fig:contract-lang}) provides
operators to compose relations. Thus, the syntax $(R_1 \cap R_2)(a,b)$
is equivalent to $R_1(a,b) \wedge R_2(a,b)$. The last line of the
above contract says that if a \cf{deposit} or a \cf{withdraw}
operation precedes a \cf{getBalance} operation in session order, and
is applied on the same object as the \cf{getBalance} operation, then
it must be the case that the \cf{getBalance} operation witnesses the
effects of the preceding operations.

Since there are no restrictions on when and how a \cf{deposit}
operation can execute, its contract is simply $\small \true$.

@@

How might one go about ensuring that any execution adheres to the contract? The
challenge is that a contract provides a declarative (axiomatic) specification
of an execution, while what is required is is an operational procedure for
\emph{enforcing} its implicit constraints. One strategy would be to execute
operations speculatively.  Here, operations are tentatively applied as they are
received from the client or other replicas. We can maintain a runtime
manifestation of executions, and check well-formedness conditions at runtime,
rolling back executions if they are ill-formed. However, the overhead of state
maintenance and the complexity of user-defined contracts make this technique
infeasible in practice.

>>>>

Consider how we might implement a highly available bank account on top of an
eventually consistent data store, with the \emph{integrity} constraint that the
balance should never drop below 0. We begin by implementing a bank
account replicated data type (RDT) in \name, and then describe the mechanisms
to obtain the desired correctness guarantees.

@@

A key novelty in \name is that it allows the addition of new RDTs to the store,
greatly improving the programmability of data stores. The programmer no
longer has to coerce the application logic to utilize the store provided data
types. In addition, \name treats the convergence semantics of the data type
separately from the consistency properties. This separation of concerns allows
the programmer to \emph{operationally} reason about conflict resolution, while
permitting \emph{declarative} reasoning for consistency. As mentioned in the
previous section, RDTs are described in terms of operations on a RDT object,
where each operation is described as a reduction over the state of the object
in terms of a list of effects. Let us assume that the bank account object provides
three operations: \cf{deposit}, \cf{withdraw} and \cf{getBalance}, with the
assumption that the withdraw fails if the account has insufficient balance.
Every operation in \name is of the following type, written in Haskell syntax:

@@

\noindent It takes a list of effects (the \emph{context} for the operation),
and an input argument, and returns a result along with an optional effect. Read-only
operations do not return any effect. If an effect is returned, it is added to
the state of the object at the current replica, and asynchronously sent to
other replicas. The implementation of the bank account operations in \name is
given in Figure~\ref{fig:ex}:

\begin{figure}
\begin{codehaskell}
data Acc = Deposit_ Int | Withdraw_ Int | GetBalance_

@@

withdraw :: [Acc] -> Int -> (Bool, Maybe Acc)
withdraw ctxt v =
	if sel1 $ getBalance ctxt () >= v
  then (True, Just $ Withdraw_ v)
	else (False, Nothing)
\end{codehaskell}
\label{Definition of a bank account expressed in Quelea.}
\label{fig:ex}
\end{figure}

@@

\name helps facilitate the mapping of operations to appropriate consistency
levels by letting the programmer declare application-level consistency
constraints as \emph{contracts} (Fig~\ref{fig:contract-lang}) that axiomatically
specify the set of allowed executions involving this operation. In the case of
the bank account, any execution that does not exhibit the anomalies
described in the previous section is a \emph{well-formed} execution on the bank
account object.  By specifying the set of legal executions for each data type in
terms of a trace of operation invocations on that type, \name\ ensures that all
executions over that type are well-formed.

@@

\noindent Here, $\cureff$ stands for the effect emitted by the \cf{withdraw} operation.
The syntax $a:\rcf{withdraw}$ states that $a$ is an effect  emitted
by a \cf{withdraw} operation i.e., $\oper{a}{\rcf{withdraw}}$ holds.  The
contract specifies that if the current operation emits an effect $\cureff$,
then for any operation $a$ which was emitted by a \cf{withdraw} operation, it
is the case that $a = \cureff$ or $a$ is visible to $\cureff$, or vice versa.
Any execution on a bank account object that preserves the above contract for a
\cf{withdraw} operation is said to be derived from a correct implementation of
\cf{withdraw}.

@@

Our specification language provides operators to compose relations.  Thus,
the syntax $(R_1 \cap R_2)(a,b)$ is equivalent to $R_1(a,b) \wedge
R_2(a,b)$. The last line of the above contract says that if a \cf{deposit}
or a \cf{withdraw} operation precedes a \cf{getBalance} operation in session
order, and is applied on the same object as the \cf{getBalance} operation,
then it must be the case that the \cf{getBalance} operation witnesses the
effects of the preceding operations. Since there are no restrictions on when
or how a \cf{deposit} operation can execute, its contract is simply $\small
\true$.

@@

How might one go about ensuring that an execution adheres to a contract? The
challenge is that a contract provides a declarative (axiomatic)
specification of an execution, while what is required is is an operational
procedure for \emph{enforcing} its implicit constraints. One strategy would
be to execute operations speculatively.  Here, operations are tentatively
applied as they are received from the client or other replicas. We can
maintain a runtime manifestation of executions, and check well-formedness
conditions at runtime, rolling back executions if they are
ill-formed. However, the overhead of state maintenance and the complexity of
user-defined contracts is likely to make this technique infeasible in
practice.

### 2014-10-30 transactions.tex
edits

While contracts on individual operations offer the programmer object-level
declarative reasoning, real-world scenarios often involve operations that span
multiple objects. In order to address this problem, several recent
systems~\cite{COPS,BurckhardtESOP,BailisHAT} have proposed a variety of
transactions, with varying semantics, in order to compose operations on
multiple objects. However, given that classical transaction models such as
serializability~\cite{} and snapshot isolation~\cite{} require inter-replica
coordination, these systems espouse \emph{coordination-free transactions} that
remain available under network partitions, but only provide weaker isolation
guarantees. Coordination-free transactions have intricate consistency semantics
and widely varying runtime overheads. As with operation-level consistency, the
onus is on the programmer to pick the correct transaction kind, the choice of
which is complicated by the consistency requirement of the operations it
composes.

@@

\name automates the choice of assigning the correct and most efficient
transaction isolation level. Similar to contracts on individual operations, the
programmer associates contracts with transactions, declaratively expressing its
consistency specification. We extend the contract language with a new term
under quantifier-free propositions ${\small \txnZ}~S_1~S_2$, where $S_1$ and
$S_2$ are sets of effects, and introduce a new primitive equivalence relation
$\small \sametxnZ$ that holds for effects from the same transaction. $\small
\txn{a,b}{c,d}$ is just syntactic sugar for $\small \sametxn{a}{b} ~\wedge~
\sametxn{c}{d} ~\wedge~ \neg\sametxn{a}{c}$, where $a$ and $b$ considered to
belong to the \emph{current} transaction.

We assume that operations not part of any transaction to belong to their own
unique transaction. While the transactions may have varying isolation
guarantees, we make the standard assumption that all transactions provide
atomicity. Hence, we include the following axiom in $\Delta$: $\small \forall
a,b,c.~\txn{a}{b,c} ~\wedge~ \sameobj{b}{c} ~\wedge~ \vis{b}{a} \Rightarrow
\vis{c}{a}$. The semantics of this contract is illustrated represented in
Figure~\ref{fig:txn_atomicity}.

@@

In order to illustrate the utility of declarative reasoning for transactions,
let us extend our running bank account example with use two accounts (objects)
-- current ($c$) and savings ($s$). Each account provides operations
\cf{withdraw}, \cf{deposit} and \cf{getBalance}, with the same contracts as
defined previously. We consider two transactions -- \cf{save(amt)}, which
transfers \cf{amt} from current to savings, and \cf{totalBalance} returns the
sum of the balances of individual accounts. The \name code for the transactions
is given below:
\vspace{-1em}

>>>>

While contracts on individual operations offer the programmer object-level
declarative reasoning, real-world scenarios often involve operations that
span multiple objects. In order to address this problem, several recent
systems~\cite{COPS,BurckhardtESOP,BailisHAT} have proposed a variety of
transactions, with varying semantics, in order to compose operations on
multiple objects. However, given that classical transaction models such as
serializability~\cite{} and snapshot isolation~\cite{} require inter-replica
coordination, these systems espouse \emph{coordination-free transactions}
that remain available under network partitions, but only provide weaker
isolation guarantees. Coordination-free transactions have intricate
consistency semantics and widely varying runtime overheads. As with
operation-level consistency, the onus is on the programmer to pick the
correct transaction kind; this selection is complicated by operation
consistency requirements.

@@

\name automates the choice of assigning the correct and most efficient
transaction isolation level. Similar to contracts on individual operations, the
programmer associates contracts with transactions, declaratively expressing its
consistency specification. We extend the contract language with a new term
under quantifier-free propositions - ${\small \txnZ}~S_1~S_2$, where $S_1$ and
$S_2$ are sets of effects, and introduce a new primitive equivalence relation
$\small \sametxnZ$ that holds for effects from the same transaction. $\small
\txn{a,b}{c,d}$ is just syntactic sugar for $\small \sametxn{a}{b} ~\wedge~
\sametxn{c}{d} ~\wedge~ \neg\sametxn{a}{c}$, where $a$ and $b$ considered to
belong to the \emph{current} transaction.

We assume that operations not part of any transaction belong to their own
unique transaction. While transactions may have varying isolation
guarantees, we make the standard assumption that all transactions provide
atomicity. Hence, we include the following axiom in $\Delta$: $\small \forall
a,b,c.~\txn{a}{b,c} ~\wedge~ \sameobj{b}{c} ~\wedge~ \vis{b}{a} \Rightarrow
\vis{c}{a}$. The semantics of this contract is illustrated represented in
Figure~\ref{fig:txn_atomicity}.

@@

In order to illustrate the utility of declarative reasoning for transactions,
let us extend our running bank account example to use two accounts (objects)
-- current ($c$) and savings ($s$). Each account provides operations
\cf{withdraw}, \cf{deposit} and \cf{getBalance}, with the same contracts as
defined previously. We consider two transactions -- \cf{save(amt)}, which
transfers \cf{amt} from current to savings, and \cf{totalBalance}, which returns the
sum of the balances of individual accounts. The \name code for the transactions
is given below:
\vspace{-1em}
